rgbif: R client for working with GBIF species occurrence data

### 2015-10-06 manuscript.Rmd
more intro text, rebuilt pdf

Data on individuals, including which species, and where they're found,
can be used for a large number of research questions. In addition,
the pool of questions we can answer becomes much larger with more
and better data. In addition to wide utility, this data is important
for conservation. Biodiversity loss is one of the greatest challenges
of our time [@Pimm_2014]. Some have called this the sixth great mass
extinction [@Ceballos_2015]. Given this challenge there is a great
need for data on specimen records, whether collected from live sightings
in the field or specimens in museums.

There are many online services that collect and maintain specimen records.
However, Global Biodiversity Information Facility (hereafter, GBIF,
http://www.gbif.org/) is the largest collection of biodiversity
records globally, currently with 580 million records, 1.6 million
taxa, 15,000 datasets from 770 publishers (figures collected on 2015-10-04).
Many large biodiversity warehouses such as iNaturalist, VertNet, and USGS's
BISON all feed into GBIF.

Herein, we describe a library (rgbif [@rgbif]) for working with
GBIF data in the R programming environment [@R]. R is an extremely
widely used language in academia, and in non-profit and private sectors.
Importantly, R makes it easy to do all of the steps of the
research process, including data management, data manipulation
and cleaning, statistics, and vizualization. Thus, an R client
for getting GBIF data is a powerful tool for reproducible research.

>>>>

Data on individuals, including which species, and where they're found,
can be used for a large number of research questions. Biodiversity
records have been used for a suite of other use cases:
validating habitat suitability models with real occurrence data
[@Ficetola_2014]; ancestral range reconstruction [@Ferretti_2015;
@Mendoza_2015]; development of invasive species watch lists
[@Faulkner_2014]; evaluate risk of invasive species spread
[@Di_Febbraro_2013]; and effects of climate change on future
biodiversity [@Brown_2015].

In addition to wide utility, this data is important for conservation.
Biodiversity loss is one of the greatest challenges of our time
[@Pimm_2014]. Some have called this the sixth great mass
extinction [@Ceballos_2015]. Given this challenge there is a great
need for data on specimen records, whether collected from live sightings
in the field or specimens in museums.

There are many online services that collect and maintain specimen records.
However, Global Biodiversity Information Facility (hereafter, GBIF,
http://www.gbif.org/) is the largest collection of biodiversity
records globally, currently with 580 million records, 1.6 million
taxa, 15,000 datasets from 770 publishers (figures collected on 2015-10-04).
Many large biodiversity warehouses such as iNaturalist
(http://www.inaturalist.org/), VertNet (http://vertnet.org/), and USGS's
Biodiversity Information Serving Our Nation (BISON; http://bison.usgs.ornl.gov/)
all feed into GBIF.

Herein, we describe the `rgbif` library [@rgbif] for working with
GBIF data in the R programming environment [@R]. R is a widely used
language in academia, and in non-profit and private sectors.
Importantly, R makes it easy to do all of the steps of the
research process, including data management, data manipulation
and cleaning, statistics, and vizualization. Thus, an R client
for getting GBIF data is a powerful tool to facilitate
reproducible research.

### 2015-10-06 manuscript.tex
more intro text, rebuilt pdf

Data on individuals, including which species, and where they're found,
can be used for a large number of research questions. In addition, the
pool of questions we can answer becomes much larger with more and better
data. In addition to wide utility, this data is important for
conservation. Biodiversity loss is one of the greatest challenges of our
time (Pimm et al., 2014). Some have called this the sixth great mass
extinction (Ceballos et al., 2015). Given this challenge there is a
great need for data on specimen records, whether collected from live
sightings in the field or specimens in museums.

There are many online services that collect and maintain specimen
records. However, Global Biodiversity Information Facility (hereafter,
GBIF, \url{http://www.gbif.org/}) is the largest collection of
biodiversity records globally, currently with 580 million records, 1.6
million taxa, 15,000 datasets from 770 publishers (figures collected on
2015-10-04). Many large biodiversity warehouses such as iNaturalist,
VertNet, and USGS's BISON all feed into GBIF.

Herein, we describe a library (rgbif (Chamberlain et al.)) for working
with GBIF data in the R programming environment (R Core Team, 2014). R
is an extremely widely used language in academia, and in non-profit and
private sectors. Importantly, R makes it easy to do all of the steps of
the research process, including data management, data manipulation and
cleaning, statistics, and vizualization. Thus, an R client for getting
GBIF data is a powerful tool for reproducible research.

@@

@@

>>>>

Data on individuals, including which species, and where they're found,
can be used for a large number of research questions. Biodiversity
records have been used for a suite of other use cases: validating
habitat suitability models with real occurrence data (Ficetola et al.,
2014); ancestral range reconstruction (Ferretti et al., 2015; Mar{í}a
Mendoza et al., 2015); development of invasive species watch lists
(Faulkner et al., 2014); evaluate risk of invasive species spread
(Febbraro et al., 2013); and effects of climate change on future
biodiversity (Brown et al., 2015).

In addition to wide utility, this data is important for conservation.
Biodiversity loss is one of the greatest challenges of our time (Pimm et
al., 2014). Some have called this the sixth great mass extinction
(Ceballos et al., 2015). Given this challenge there is a great need for
data on specimen records, whether collected from live sightings in the
field or specimens in museums.

There are many online services that collect and maintain specimen
records. However, Global Biodiversity Information Facility (hereafter,
GBIF, \url{http://www.gbif.org/}) is the largest collection of
biodiversity records globally, currently with 580 million records, 1.6
million taxa, 15,000 datasets from 770 publishers (figures collected on
2015-10-04). Many large biodiversity warehouses such as iNaturalist
(\url{http://www.inaturalist.org/}), VertNet
(\url{http://vertnet.org/}), and USGS's Biodiversity Information Serving
Our Nation (BISON; \url{http://bison.usgs.ornl.gov/}) all feed into
GBIF.

Herein, we describe the \texttt{rgbif} library (Chamberlain et al.) for
working with GBIF data in the R programming environment (R Core Team,
2014). R is a widely used language in academia, and in non-profit and
private sectors. Importantly, R makes it easy to do all of the steps of
the research process, including data management, data manipulation and
cleaning, statistics, and vizualization. Thus, an R client for getting
GBIF data is a powerful tool to facilitate reproducible research.

@@

Brown KA., Parks KE., Bethell CA., Johnson SE., Mulligan M. 2015.
Predicting plant diversity patterns in madagascar: Understanding the
effects of climate and land cover change in a biodiversity hotspot.
\emph{PLOS ONE} 10:e0122721.

@@

Faulkner KT., Robertson MP., Rouget M., Wilson JR. 2014. A simple, rapid
methodology for developing invasive species watch lists.
\emph{Biological Conservation} 179:25--32.

Febbraro MD., Lurz PWW., Genovesi P., Maiorano L., Girardello M.,
Bertolino S. 2013. The use of climatic niches in screening procedures
for introduced species to evaluate risk of spread: A case with the
american eastern grey squirrel. \emph{PLoS ONE} 8:e66559.

Ferretti F., Verd GM., Seret B., {Š}prem JS., Micheli F. 2015. Falling
through the cracks: The fading history of a large iconic predator.
\emph{Fish and Fisheries}:n/a--n/a.

Ficetola GF., Rondinini C., Bonardi A., Baisero D., Padoa-Schioppa E.
2014. Habitat availability for amphibians and extinction threat: A
global analysis. \emph{Diversity and Distributions} 21:302--311.

Mar{í}a Mendoza., Ospina OE., C{á}rdenas-Henao H., Garc{í}a-R JC. 2015.
A likelihood inference of historical biogeography in the world's most
diverse terrestrial vertebrate genus: Diversification of
direct-developing frogs (craugastoridae: Pristimantis) across the
neotropics. \emph{Molecular Phylogenetics and Evolution} 85:50--58.
